Concrete Semantics for Pushdown Analysis:\\The Essence of Summarization

### 2013-12-09 content.tex
Tightening wording and split out some text to 'tech report' version using an iftr macro.

%%
There is a systematic process for transforming off-the-shelf programming language semantics into a form amenable to \emph{regular} analysis that has been widely applied with great success to production programming languages~\citet{dvanhorn:VanHorn2010Abstracting} (a technique called abstracting abstract machines, or AAM).
%
A contribution of this paper is a systematic process to construct \emph{pushdown} analyses of programming languages, due to the precision (and often performance) benefits.
%

@@

CFA2 is an online version of a technique called ``summarization'' from Sharir and Pnueli \citep[Chapter 7]{local:muchnick:jones:flow-analysis:1981}, which is synonymous with the $\epsilon$-closure graph that PDCFA constructs.
%
The online quality of the algorithm is what makes it suitable for higher-order languages.
%
Summarization algorithms need not be restricted to languages with well-bracketed calls and returns.
%
We can adopt the technique for higher precision in the common case but still handle difficult cases such as first-class control.
%
This was shown for the \rackett{call-with-current-continuation} (a.k.a. \rackett{call/cc}) operator in~\citet{ianjohnson:Vardoulakis2011Pushdown}.
%
This work illuminated the fact that we can harness the enhanced technology of pushdown analyses in non-pushdown models of computation.
%
Doing this sacrifices call/return matching in the general case, but in practice the precision is much better than the alternative regular model that, say, 0-CFA would provide.
%%

%%
A downside of the work providing \rackett{call/cc} is that it was an algorithmic change to the already complicated CFA2 --- there was no recipe for how to do this for one's favorite control operator.
%
This paper seeks to do just that with an operational view of what summarization is, in essence.
%
In other words, we give a concrete semantics to the tricks that the analyses in the literature use in the abstract, and maintain the original meaning of the language.
%
We also give intuitive analogies to well-established ideas and techniques so that the working semanticist can write a pushdown analysis for her language.
%
In order to demonstrate the applicability of this viewpoint, we show a new analysis for a language with composable control.
%
All of the semantics modeled in this paper are implemented in full detail in PLT Redex~\citep{ianjohnson:Felleisen:2009:SEP:1795772} and available online.\footnote{\url{http://github.com/ianj/concrete-summaries}}

There are common underpinnings of PDCFA and CFA2 that can be embodied as concrete ``summarization'' machinery in the programming language semantics: 
\begin{enumerate}
\item{bound continuations with with delimiters that specify the exact context in which any continuation replacement would lead to the same result, and}
\item{memoize at these delimiters}
\end{enumerate}

%%
The first of these two might seem unfamiliar, but it is just identifying what is needed for sound memoization, and we delimit the continuations in order to mark where to memoize results.
%
In the case of a CESK machine, where the store is relevant to the evaluation of functions, a context is just the CES portion since stacks don't matter.
%
For stack inspecting analyses, a context is CES plus the inspected portion of K.
%
As soon as we have a value that such a context evaluates to, all subsequent visits to that context can jump straight to the memoized result(s).
%%

@@

PDCFA does not have some orthogonal semantic components that CFA2 features to improve precision and is thus the simpler of the two.
%
The key insight of PDCFA is to notice in languages without stack-capturing features, the continuation is only modified a frame at a time.
%
If the state space without the stack is finite (and it can be made finite with an approximation ala AAM), then the model falls directly into the realm of pushdown systems.
%
They thus recast the problem in terms of pushdown systems.
%
The whole of their machinery is then computing the pushdown system on-the-fly, only considering states reachable from the initial state.
%
They call the main data-structure for this a Dyck state graph.
%
They additionally compute an $\epsilon$-closure graph to match nodes that push frames with later nodes following the pop of that frame, which keeps edges between nodes that have a net-zero stack change path between them.
%

@@

%%
The $\Addr$ space is what controls the precision of the model.
%
For a concrete semantics, we require the allocation meta-function $\alloc :
\State \to \Addr$ to return fresh addresses ($\alloc(\tpl{(\mexpr, \menv), \mstore, \mkont}) \notin \dom(\mstore)$), but any choice of address is sound~\citep{dvanhorn:Might2009Posteriori}.
%
If $\alloc$ only uses addresses from a finite subset of $\Addr$, then the state space without continuations is finite, and the space of continuation frames is finite.
%
A finite allocation strategy makes the continuation the only unbounded data structure in the machine, which follows a stack discipline with a finite ``stack alphabet'' --- thus the pushdown system interpretation is apt.
%
The $\alloc$ meta-function along with the commitment to having no recursive data-structures is the key to the AAM technique.
%
All recursion can be expressed by indirecting through addresses into a store.
%
We borrow this technique on top of our own.

@@

\paragraph{The role of summaries:} notice the additional $\Memo$ component.
%
A ``summary edge'' in CFA2 or equivalently an $\epsilon$-edge in PDCFA is an edge from the source of a push edge to the target of the matching pop edge.
%
``Matching'' here means there is a path through machine reductions that don't change the stack, or through summary edges.
%
In our model, we only ``push'' when we call a function, and we only ``pop'' when we return with a value.
%
There is an analogy to something more operational: summary edges embody memoization (see the last rule in figure \ref{fig:summary-semantics}).
%
Instead of following an entire path through a call to return a value, we simply jump from the call to the return with the result of the call, with the resulting context changes---in this case only the store (the second case of rule 4).
%%

\begin{figure}
  \centering
  $\mstate \stepto \mstate' \text{ where } \maddr = \alloc(\mstate)$ \\
  \begin{tabular}{r|l}
    \hline
% variable lookup
    $\tpl{(\svar\mvar, \menv), \mstore, \mkont, \mktab, \mmemo}$
    &
    $\tpl{\mval,\mstore,\mkont, \mktab, \mmemo}$ if $\mval \in \mstore(\menv(\mvar))$
    \\
% application
    $\tpl{(\sapp{\mexpri0}{\mexpri1}, \menv), \mstore, \mkont, \mktab, \mmemo}$
    &
    $\tpl{(\mexpri0, \menv), \mstore, \kcons{\kar{\mexpri1,\menv}}{\mkont}, \mktab, \mmemo}$
    \\
% argument evaluation
    $\tpl{\mval, \mstore, \kcons{\kar{\mexpr,\menv}}{\mkont}, \mktab, \mmemo}$
    &
    $\tpl{(\mexpr, \menv), \mstore, \kcons{\kfn{\mval}}{\mkont}, \mktab, \mmemo}$
    \\
% function call
    $\tpl{\mval,\mstore,\kcons{\kfn{\vclo{\slam{\mvar}{\mexpr}}{\menv}}}{\mkont},\mktab,\mmemo}$
    & % actually do call,
    $\tpl{\mpoint,
          \mstore',
          \krt{\mpoint, \mstore'},
          \mktab',
          \mmemo}$
     \\ & \quad if $(\mpoint, \mstore') \notin \dom(\mmemo)$
\\
    & or \\
    & $\tpl{\mval_\mathit{result},
            \mstore'',
            \mkont,
            \mktab',
            \mmemo}$
    \\ & \quad if $(\mval_\mathit{result},\mstore'') \in \mmemo(\mpoint,\mstore')$
    \\ % or, lookup in table
%    & $\tpl{\mval, \mstore', \mkont,\mktab,\mmemo}$ if $\mval \in \mmemo(\mpoint, \mstore')$ \\    
    where & $\mpoint = (\mexpr, \extm{\menv}{\mvar}{\maddr})$ \\
          & $\mstore' = \joinone{\mstore}{\maddr}{\mval}$ \\
          & $\mktab' = \joinone{\mktab}{(\mpoint, \mstore')}{\mkont}$
    \\
% return
    $\tpl{\mval, \mstore, \krt{\mpoint,\mstore'}, \mktab, \mmemo}$
    &
    $\tpl{\mval, \mstore, \mkont, \mktab, \joinone{\mmemo}{(\mpoint, \mstore')}{(\mval,\mstore)}}$
    \\ & \quad if $\mkont \in \mktab(\mpoint, \mstore')$
  \end{tabular}
  \caption{The summarizing tabular stack machine}
  \label{fig:summary-semantics}
\end{figure}

This semantics is only correct for fresh allocation strategies, which in turn make the reduction relation deterministic.
%
As it is, the CESK machine simulates the tabular machine, but not the other way around;
%
if we use a finite allocation strategy, then we don't explore some non-deterministic paths that lead to different results for a function call.
%
We thus explore all paths and share the k-table and memo tables between all paths.

To turn this semantics into PDCFA's algorithm for constructing a Dyck state graph, given the appropriate $\alloc$, we apply a widening operator to make $\mstore$, $\mkont$, and $\mmemo$ shared amongst all states (meaning states then become represented without these components), in what is then called a $\System$. We use a meta-function $\wn$, ``wide to narrow,'' to shift between the two representations of states.
%
We show only the memo and continuation table widening, since store widening is standard.
%

{
\setlength{\abovedisplayskip}{0pt}
\setlength{\belowdisplayskip}{4pt}
\setlength{\abovedisplayshortskip}{0pt}
\setlength{\belowdisplayshortskip}{8pt}
\begin{align*}
  \mastate \in \sa{State} &= (\Expr \times \Env) \times \Store \times \Kont \\
  \System &= \wp(\sa{State}) \times \wp(\sa{State}) \times \KTab \times \Memo \\
  {\mathcal F} &: \sa{\State} \to \System \to \System \\
  {\mathcal F}(\mastate_0)(S, \emptyset, \mktab, \mmemo) &= (S, \emptyset, \mktab, \mmemo) \\
  {\mathcal F}(\mastate_0)(S, F\cup\set{\mastate}, \mktab, \mmemo) &= (S \cup S', F\cup F', \mktab', \mmemo') \\
  \text{where } I &= \setbuild{\mstate'}{\wn(\mastate,\mktab,\mmemo) \stepto \mstate'} \\
                \wn(\tpl{(\mexpr,\menv),\mstore,\mkont},\mktab,\mmemo)
                   &= \tpl{(\mexpr,\menv),\mstore,\mkont,\mktab,\mmemo}
\end{align*}
\begin{minipage}{.55\linewidth}
  \begin{align*}
    \mktab' &=  \bigsqcup\setbuild{\mktab'}{\wn(\_, \mktab', \_) \in I} \\
    \mmemo' &= \bigsqcup\setbuild{\mmemo'}{\wn(\_, \_, \mmemo') \in I}
  \end{align*}
\end{minipage}
\begin{minipage}{.40\linewidth}
  \begin{align*}
    S' &= \setbuild{\mastate'}{\wn(\mastate', \_, \_) \in I} \cup \set{\mastate_0}\\
    F' &= S' \setminus S
  \end{align*}
\end{minipage}
}
%%

@@

%%
We should expect that a pushdown analysis would predict this evaluates to true, and there are no loops in the program.
%
0CFA~\citep{dvanhorn:Shivers:1991:CFA} claims there is a loop from the second call of \texttt{id} to the first, and thusly predicts this program evaluates to true or false.
%
PDCFA claims there is no loop, and depending on the implementation, that the result is true.
%
The example evaluation is in \autoref{fig:ex-eval}.
%
We maintain enough context to distinguish the return points of \texttt{id} to not rebind \texttt{y} to 1.
%
When determining control flow through the expression, we consult $\mktab$ to continue past function returns, so there is no confusion about a back edge from the second call to \texttt{id}.
%

\begin{figure}
  \centering
    \begin{minipage}{0.55\linewidth}
      \begin{center}
        $\begin{array}{l}\includegraphics[scale=0.5]{example1}\end{array}$ \\
        $\Downarrow$ \\
        $\begin{array}{l}\includegraphics[scale=0.5]{example2}\end{array}$ \\
        $\Downarrow$ \\
        $\begin{array}{l}\includegraphics[scale=0.5]{example3}\end{array}$
      \end{center}
    \end{minipage}
    \begin{minipage}{0.43\linewidth}
      \begin{center} {\small
        \begin{tabular}{|l|c|c|}
          \hline
          \# & $\mstore$ & $\mkont$ \\
          \hline
          0 & $[{\tt id} \mapsto \set{(\slam{\mvar}{\mvar},\iota)}]$ & $\mkont_0$ \\
          1 & $\mstore_0[{\tt \mvar} \mapsto \set{0}]$ & $\krt{\mctx_0}$ \\
          2 & $\mstore_1[{\tt y} \mapsto \set{0}]$ & $\mathbf{lt}({\tt z}, {\tt (}\le\ {\tt y}\ {\tt z)}, \iota)$ \\
          3 & $\mstore_2[{\tt x} \mapsto \set{0,1}]$ & $\krt{\mctx_1}$ \\
          4 & $\mstore_3[{\tt z} \mapsto \set{0,1}]$ & $\kmt$ \\
          \hline
        \end{tabular}} \\[10pt]
      {\small
        \begin{tabular}{|l|c|}
          \hline
          \# & $\mctx$ \\
          \hline
          0  & $(({\tt x}, \iota), \mstore_1)$ \\
          1  & $(({\tt x}, \iota), \mstore_3)$ \\
          \hline
        \end{tabular}}
      \\[10pt]
      {\small
        \begin{tabular}{|l|c|c|c|}
          \hline
          \# & $\mktab$ & $\mmemo$ \\
          \hline
          0  & $[\mctx_0 \mapsto \set{\mkont_0}]$ & $[\mctx_0 \mapsto \set{0}]$ \\
          1  & $\mktab_0[\mctx_1 \mapsto \set{\mkont_2}]$ & $\mmemo_0[\mctx_1 \mapsto \set{0,1}]$ \\
          \hline
        \end{tabular}}    
    \end{center}
  \end{minipage}
Where $\mkont_0 = \mathbf{lt}({\tt y}, {\tt (let\ ([z\ (id\ 1)])\ (}\le\ {\tt y}\ {\tt z))}, \iota)$
\caption{Example analysis evaluation}
\label{fig:ex-eval}
\end{figure}
% %% FIXME stores in memo-table
% \begin{enumerate}
% \item{\texttt{(id 0)} steps to \texttt{x} at $\krt{((\texttt{x}, [\texttt{x}\mapsto\texttt{x}]), \mstore_0)}$ with $\mstore_1 =
%     \mstore_0[\texttt{x} \mapsto \set{0}]$ and
%     (let $\mctx_1 = ((\texttt{x},[\texttt{x}\mapsto\texttt{x}]),\mstore_1)$,
%          $\mkont_1 = \klet{\texttt{y}, \texttt{(let ([z (id 1)]) ($\le$ y z))}, [\texttt{id}\mapsto{\texttt{id}}]}$)
%     $\mktab_1 = [\mctx_0 \mapsto \set{\mkont_0}]$}
% \item{\texttt{0} at $\krt{\mctx_1}$ steps to $0$ at $\mkont_1$ and $\mmemo_1 = [\mctx_1 \mapsto \set{0}]$}
% \item{\texttt{0} at $\mkont_1$ steps to \texttt{(let ([z (id 1)]) ($\le$ y z))} and $\mstore_2 = \mstore_1[\texttt{y} \mapsto \set{0}]$}
% \item{\texttt{(let ([z (id 1)]) ($\le$ y z))} steps to \texttt{(id 1)} at \\
%       $\klet{\texttt{z}, \texttt{($\le$ y z)}, [\texttt{id} \mapsto \texttt{id}, \texttt{y} \mapsto \texttt{y}]}$
%       (call this $\mkont_2$).}
% \item{\texttt{(id 1)} steps to $x$ with $\mstore_3 = \mstore_2[\texttt{x} \mapsto \set{0,1}]$ and
%       (let $\mctx_3 = ((\texttt{x},[\texttt{x}\mapsto\texttt{x}]),\mstore_3)$) $\mktab_3 =\mktab_1[\mctx_3 \mapsto \set{\mkont_2}]$.}
% \item{\texttt{0} or \texttt{1} at $\krt{\mctx_3}$ steps to \texttt{0} or \texttt{1} at $\mkont_2$ and $\mmemo_3 = \mmemo_1[\mctx_3 \mapsto \set{0,1}]$.}
% \item{\texttt{z} gets bound to $\set{0,1}$, and \texttt{($\le$ y z)} evaluates to true.}
% \end{enumerate}

\subsection{Metatheory}

@@

\begin{claim}(Correctness of Concrete Summarization)\label{thm:concrete-tabular}
  The standard semantics has a bisimulation with the tabular semantics given comparable fresh allocation strategies.
\end{claim}

Finally, we can show that the reductions we find between states paired with the shared tables define a sound and complete abstraction of the CESK machine with a comparable allocation strategy.
\begin{claim}[Correctness of PDCFA]\label{thm:global}
  Widened tabular semantics reifies a relation that has a bisimulation with the standard semantics.
\end{claim}

@@

%%
CFA2 also includes what they called ``transitive summaries'' to deal with tail calls.
%
We create an $\rt$ continuation at every function call.
%
This view contends with tail calls, but we can identify tail calls easily --- any call with an $\rt$ as its continuation is a tail call.
%
Tail calls are important in language implementations for space complexity reasons \citep{ianjohnson:clinger:tail-calls:1998}, but in an analysis, these concerns are less important.
%
The repeated popping of $\rt$ inserted by what otherwise were tail-calls is synonymous with CFA2's transitive summaries.

>>>>

%%
There is a systematic process for transforming off-the-shelf programming language semantics into a form amenable to \emph{regular} analysis that has been widely applied with great success to production programming languages (a technique called abstracting abstract machines, or AAM~\citep{dvanhorn:VanHorn2010Abstracting}).
%
A contribution of this paper is a systematic process to construct \emph{pushdown} analyses of programming languages, due to the precision (and often performance) benefits.
%

@@

CFA2 is an online version of a technique called ``summarization'' from Sharir and Pnueli \citep[Chapter 7]{local:muchnick:jones:flow-analysis:1981}, which is synonymous with the $\epsilon$-closure graph that PDCFA constructs.
%
The online quality of the algorithm is what makes it suitable for higher-order languages.
%
Summarization algorithms need not be restricted to languages with well-bracketed calls and returns.
%
We can adopt the technique for higher precision in the common case but still handle difficult cases such as first-class control.
%
This was shown for the \rackett{call-with-current-continuation} (or \rackett{call/cc}) operator in~\citet{ianjohnson:Vardoulakis2011Pushdown}.
%
This work illuminated the fact that we can harness the enhanced technology of pushdown analyses in non-pushdown models of computation.
%
Doing this sacrifices call/return matching in the general case, but in practice the precision is much better than the alternative regular model that, say, 0-CFA would provide.
%%

%%
This paper extends the work providing \rackett{call/cc} by giving an operational view of what summarization is, in essence.
%
In other words, we give a concrete semantics to the algorithmic techniques that the analyses in the literature use in the abstract, and maintain the original meaning of the language.
%
We also give intuitive analogies to well-established ideas and techniques so that the working semanticist can write a pushdown analysis for their language.
%
In order to demonstrate the applicability of this viewpoint, we show a new analysis for a language with composable control.
%
All of the semantics modeled in this paper are implemented in full detail in PLT Redex~\citep{ianjohnson:Felleisen:2009:SEP:1795772} and available online.\footnote{\url{http://github.com/ianj/concrete-summaries}}

There are common underpinnings of PDCFA and CFA2 that can be embodied as concrete ``summarization'' machinery in the programming language semantics: 
\begin{enumerate}
\item{memoize calls with a ``return'' stack frame that tracks calling context, and}
\item{store stacks that reach calls in order to continue to them after memoizing}
\end{enumerate}

%%
The first of these is a semantic approach to memoizing that is a straight-forward analog to the programmatic approach.
%
A memoizing program will call a function in a non-tail context in order to obtain the result, store the result to an ``enough context to obtain the same result'' key in a table (in a pure setting, just the function and arguments), and continue with the result.
%
The semantics we use is functionally equivalent, only we have a special stack frame that acts as the non-tail context that additionally stores results in a table that is part of the machine state.
%
In the case of a baseline abstract machine for AAM, a CESK machine~\citep{dvanhorn:Felleisen1987Calculus}, the store is relevant to the evaluation of functions.
%
As such, the CES portion of a state suffices as ``enough context.''
%
For a semantics that inspects the stack, ``enough context'' is CES plus the inspected portion of K.
%%

%%
The second piece of machinery serves two purposes:
%
(1) the state space contains all function callsites to project out a control-flow graph, and %
(2) the worklist semantics of execution makes it possible to appeal to an incomplete memo table, so the saved stacks allow new results to flow to calls that used incomplete memo information.
%
Consider a non-deterministic semantics where a function call reaches a result and memoizes, a second call uses the memo table (storing its stack), and then the original call reaches another result (since the worklist allows for such parallel reduction).
%
The second result had better also flow to the second call to maintain soundness!
%%

@@

PDCFA does not have some orthogonal semantic components that CFA2 features to improve precision and is thus the simpler of the two.
%
The key insight of PDCFA is to notice in languages without stack-capturing features, the continuation is only modified a frame at a time.
%
If the state space without the stack is finite (and it can be made finite with an approximation ala AAM), then the model is naturally recast into the realm of pushdown systems.
%
The whole of their machinery is then computing the pushdown system on-the-fly, only considering states reachable from the initial state.
%
A key component is computing the $\epsilon$-closure graph to match nodes that push frames with later nodes following the pop of that frame.
%
This stores all edges between nodes that have a net-zero stack change path between them.
%

@@

%%
The $\Addr$ space is what controls the precision of the model.
%
For a concrete semantics, we require the allocation meta-function $\alloc :
\State \to \Addr$ to return fresh addresses ($\alloc(\mstate) \notin \dom(\mstore)$), but any choice of address is sound~\citep{dvanhorn:Might2009Posteriori}.
%
If $\alloc$ only uses addresses from a finite subset of $\Addr$, then the state space without continuations is finite, and the space of continuation frames is finite.
%
A finite allocation strategy makes the continuation the only unbounded data structure in the machine, which follows a stack discipline with a finite ``stack alphabet'' --- thus the pushdown system interpretation is apt.
%
The $\alloc$ meta-function along with the commitment to having no recursive data-structures is the key to the AAM technique.
%
All recursion can be expressed by indirecting through addresses into a store.
%
We borrow this technique on top of our own.

@@

\paragraph{The role of summaries:} notice the additional $\Memo$ component.
%
A ``summary edge'' in CFA2 or equivalently an $\epsilon$-edge in PDCFA is an edge from the source of a push edge to the target of the matching pop edge.
%
``Matching'' here means there is a path through machine reductions that don't change the stack, or through summary edges.
%
In our model, we only ``push'' when we call a function, and we only ``pop'' when we return with a value.
%
There is an analogy to something more operational: summary edges embody memoization.
%
Instead of following an entire path through a call to return a value, we simply jump from the call to the return with the result of the call, with the resulting context changes---in this case only the store [memo lookup].
%%

\begin{figure}
  \centering
  $\mstate \stepto \mstate' \text{ where } \maddr = \alloc(\mstate)$ \\
  \begin{tabular}{r|ll}
    \hline
% variable lookup
    $\tpl{(\svar\mvar, \menv), \mstore, \mkont, \mktab, \mmemo}$
    &
    $\tpl{\mval,\mstore,\mkont, \mktab, \mmemo}$ if $\mval \in \mstore(\menv(\mvar))$
    &
    [variable lookup]
    \\
% application
    $\tpl{(\sapp{\mexpri0}{\mexpri1}, \menv), \mstore, \mkont, \mktab, \mmemo}$
    &
    $\tpl{(\mexpri0, \menv), \mstore, \kcons{\kar{\mexpri1,\menv}}{\mkont}, \mktab, \mmemo}$
    &
    [function eval]
    \\
% argument evaluation
    $\tpl{\mval, \mstore, \kcons{\kar{\mexpr,\menv}}{\mkont}, \mktab, \mmemo}$
    &
    $\tpl{(\mexpr, \menv), \mstore, \kcons{\kfn{\mval}}{\mkont}, \mktab, \mmemo}$
    &
    [argument eval]
    \\
% function call
    $\tpl{\mval,\mstore,\kcons{\kfn{\vclo{\slam{\mvar}{\mexpr}}{\menv}}}{\mkont},\mktab,\mmemo}$
    & % actually do call,
    $\tpl{\mpoint,
          \mstore',
          \krt{\mpoint, \mstore'},
          \mktab',
          \mmemo}$
    &
    [un-memoized call] 
     \\ & \quad if $(\mpoint, \mstore') \notin \dom(\mmemo)$
\\
    & or \\
    & $\tpl{\mval_\mathit{result},
            \mstore'',
            \mkont,
            \mktab',
            \mmemo}$
    & [memo lookup]
    \\ & \quad if $(\mval_\mathit{result},\mstore'') \in \mmemo(\mpoint,\mstore')$
    \\ % or, lookup in table
%    & $\tpl{\mval, \mstore', \mkont,\mktab,\mmemo}$ if $\mval \in \mmemo(\mpoint, \mstore')$ \\    
    where & $\mpoint = (\mexpr, \extm{\menv}{\mvar}{\maddr})$ \\
          & $\mstore' = \joinone{\mstore}{\maddr}{\mval}$ \\
          & $\mktab' = \joinone{\mktab}{(\mpoint, \mstore')}{\mkont}$
    \\
% return
    $\tpl{\mval, \mstore, \krt{\mpoint,\mstore'}, \mktab, \mmemo}$
    &
    $\tpl{\mval, \mstore, \mkont, \mktab, \joinone{\mmemo}{(\mpoint, \mstore')}{(\mval,\mstore)}}$
    &
    [memoize/return]
    \\ & \quad if $\mkont \in \mktab(\mpoint, \mstore')$
  \end{tabular}
  \caption{The summarizing tabular stack machine}
  \label{fig:summary-semantics}
\end{figure}

This semantics is only correct for fresh allocation strategies, which in turn make the reduction relation deterministic.
%
As it is, the CESK machine simulates the tabular machine, but not the other way around;
%
if we use a finite allocation strategy, then we don't explore some non-deterministic paths that lead to different results for a function call.
%
We thus explore all paths and share the KTable and memo tables between all paths.

To turn this semantics into PDCFA's algorithm for constructing a pushdown system with only reachable states, given the appropriate $\alloc$, we apply a widening operator to make $\mkont$, and $\mmemo$ shared amongst all states (meaning states then become represented without these components), in what is then called a $\System$. We use a meta-function $\wn$, ``wide to narrow,'' to shift between the two representations of states.
%

{
\setlength{\abovedisplayskip}{0pt}
\setlength{\belowdisplayskip}{4pt}
\setlength{\abovedisplayshortskip}{0pt}
\setlength{\belowdisplayshortskip}{8pt}
\begin{align*}
  \mastate \in \sa{State} &= (\Expr \times \Env) \times \Store \times \Kont \\
  \System &= \wp(\sa{State}) \times \wp(\sa{State}) \times \KTab \times \Memo \\
  {\mathcal F} &: \sa{\State} \to \System \to \System \\
  {\mathcal F}(\mastate_0)(S, \emptyset, \mktab, \mmemo) &= (S, \emptyset, \mktab, \mmemo) \\
  {\mathcal F}(\mastate_0)(S, F\uplus\set{\mastate}, \mktab, \mmemo) &= (S \cup S', F\cup F', \mktab', \mmemo') \\
  \text{where } I &= \setbuild{\mstate'}{\wn(\mastate,\mktab,\mmemo) \stepto \mstate'} \\
                \wn(\tpl{(\mexpr,\menv),\mstore,\mkont},\mktab,\mmemo)
                   &= \tpl{(\mexpr,\menv),\mstore,\mkont,\mktab,\mmemo}
\end{align*}
\begin{minipage}{.55\linewidth}
  \begin{align*}
    \mktab' &=  \bigsqcup\setbuild{\mktab'}{\wn(\_, \mktab', \_) \in I} \\
    \mmemo' &= \bigsqcup\setbuild{\mmemo'}{\wn(\_, \_, \mmemo') \in I}
  \end{align*}
\end{minipage}
\begin{minipage}{.40\linewidth}
  \begin{align*}
    S' &= \setbuild{\mastate'}{\wn(\mastate', \_, \_) \in I} \cup \set{\mastate_0}\\
    F' &= S' \setminus S
  \end{align*}
\end{minipage}
}
%%

@@

%%
We should expect that a pushdown analysis would predict this evaluates to true, and there are no loops in the program.
%
0-CFA~\citep{dvanhorn:Shivers:1991:CFA} claims there is a loop from the second call of \texttt{id} to the first, and thusly predicts this program evaluates to true or false.
%
PDCFA claims there is no loop and that the result is true.
%
The example evaluation is in \autoref{fig:ex-eval}.
%
We maintain enough context to distinguish the return points of \texttt{id} to not rebind \texttt{y} to 1.
%
When determining control flow through the expression, we consult $\mktab$ to continue past function returns, so there is no confusion about a back edge from the second call to \texttt{id}.
%

\begin{figure}
      \begin{tabular}{rcl}
        $\begin{array}{l}\includegraphics[scale=0.5]{example1}\end{array}$
        & $\Rightarrow$ &
        $\begin{array}{l}\includegraphics[scale=0.5]{example2}\end{array}$ \\
        & & \multicolumn{1}{c}{$\Downarrow$} \\
        \begin{minipage}{0.5\linewidth}
          \begin{center}
            {\small
              \begin{tabular}{|l|c|c|c|}
                \hline
                \# & $\mktab$ & $\mmemo$ \\
                \hline
                0  & $[\mctx_0 \mapsto \set{\mkont_0}]$ & $[\mctx_0 \mapsto \set{0}]$ \\
                1  & $\mktab_0[\mctx_1 \mapsto \set{\mkont_2}]$ & $\mmemo_0[\mctx_1 \mapsto \set{0,1}]$ \\
                \hline
              \end{tabular}} \\[8pt]
            {\small
              \begin{tabular}{|l|c|}
                \hline
                \# & $\mctx$ \\
                \hline
                0  & $(({\tt x}, \iota), \mstore_1)$ \\
                1  & $(({\tt x}, \iota), \mstore_3)$ \\
                \hline
              \end{tabular}}
          \end{center}

      \end{minipage} & &
        $\begin{array}{l}\includegraphics[scale=0.5]{example3}\end{array}$
      \end{tabular}
    \begin{center} {\small
        \begin{tabular}{|l|c|c|}
          \hline
          \# & $\mstore$ & $\mkont$ \\
          \hline
          0 & $[{\tt id} \mapsto \set{(\slam{\mvar}{\mvar},\iota)}]$ & $\mkont_0$ \\
          1 & $\mstore_0[{\tt \mvar} \mapsto \set{0}]$ & $\krt{\mctx_0}$ \\
          2 & $\mstore_1[{\tt y} \mapsto \set{0}]$ & $\mathbf{lt}({\tt z}, {\tt (}\le\ {\tt y}\ {\tt z)}, \iota)$ \\
          3 & $\mstore_2[{\tt x} \mapsto \set{0,1}]$ & $\krt{\mctx_1}$ \\
          4 & $\mstore_3[{\tt z} \mapsto \set{0,1}]$ & $\kmt$ \\
          \hline
        \end{tabular}} \\[10pt]
 Where $\mkont_0 = \mathbf{lt}({\tt y}, {\tt (let\ ([z\ (id\ 1)])\ (}\le\ {\tt y}\ {\tt z))}, \iota)$
    \end{center}
\caption{Example analysis evaluation}
\label{fig:ex-eval}
\end{figure}
% %% FIXME stores in memo-table
% \begin{enumerate}
% \item{\texttt{(id 0)} steps to \texttt{x} at $\krt{((\texttt{x}, [\texttt{x}\mapsto\texttt{x}]), \mstore_0)}$ with $\mstore_1 =
%     \mstore_0[\texttt{x} \mapsto \set{0}]$ and
%     (let $\mctx_1 = ((\texttt{x},[\texttt{x}\mapsto\texttt{x}]),\mstore_1)$,
%          $\mkont_1 = \klet{\texttt{y}, \texttt{(let ([z (id 1)]) ($\le$ y z))}, [\texttt{id}\mapsto{\texttt{id}}]}$)
%     $\mktab_1 = [\mctx_0 \mapsto \set{\mkont_0}]$}
% \item{\texttt{0} at $\krt{\mctx_1}$ steps to $0$ at $\mkont_1$ and $\mmemo_1 = [\mctx_1 \mapsto \set{0}]$}
% \item{\texttt{0} at $\mkont_1$ steps to \texttt{(let ([z (id 1)]) ($\le$ y z))} and $\mstore_2 = \mstore_1[\texttt{y} \mapsto \set{0}]$}
% \item{\texttt{(let ([z (id 1)]) ($\le$ y z))} steps to \texttt{(id 1)} at \\
%       $\klet{\texttt{z}, \texttt{($\le$ y z)}, [\texttt{id} \mapsto \texttt{id}, \texttt{y} \mapsto \texttt{y}]}$
%       (call this $\mkont_2$).}
% \item{\texttt{(id 1)} steps to $x$ with $\mstore_3 = \mstore_2[\texttt{x} \mapsto \set{0,1}]$ and
%       (let $\mctx_3 = ((\texttt{x},[\texttt{x}\mapsto\texttt{x}]),\mstore_3)$) $\mktab_3 =\mktab_1[\mctx_3 \mapsto \set{\mkont_2}]$.}
% \item{\texttt{0} or \texttt{1} at $\krt{\mctx_3}$ steps to \texttt{0} or \texttt{1} at $\mkont_2$ and $\mmemo_3 = \mmemo_1[\mctx_3 \mapsto \set{0,1}]$.}
% \item{\texttt{z} gets bound to $\set{0,1}$, and \texttt{($\le$ y z)} evaluates to true.}
% \end{enumerate}

\iftr{%
\subsection{Metatheory}

@@

\begin{theorem}[Correctness of Concrete Summarization]\label{thm:concrete-tabular}
  The standard semantics has a skipping bisimulation with the tabular semantics given comparable fresh allocation strategies.
\end{theorem}
\begin{proof}[Sketch]
  The tabular semantics skipping simulates that standard semantics by appealing to the table invariants to show the long ``recomputation'' path.
  Crucially, since the memo table is only set once before being used to bypass to the result, there must be only one possible result in the standard semantics. Otherwise, the tabular semantics would miss out on the other results for subsequent calls with the same context.
  The fresh allocation strategy ensures that relation is deterministic and thus function calls only reduce to one result.

  The standard semantics skipping simulates the tabular semantics because ``recomputing'' paths always have the calling stack as its tail, so we can step forward on the invariant's inductively built path until the result.
\end{proof}

Finally, we can show that the reductions we find between states paired with the shared tables define a sound and complete abstraction of the CESK machine with a comparable allocation strategy.
}
\begin{theorem}[Correctness of PDCFA]\label{thm:global}
  With a fair state selection from the frontier, the widened tabular semantics reflects a relation that has a bisimulation with the standard semantics.
\end{theorem}
\iftr{%
\begin{proof}[Sketch]
  The invariants on the tables still hold. 
  The widening to share $\mmemo$ and $\mktab$ amongst all explored states reifies the search state into a $\sa{System}$. We can add a set of edges simply.
  If we add $\mmemo$ and $\mktab$ to all states in these edges, we get a reduction relation comparable to the unwidened tabular semantics.
  We prove that the infinite unrolling of the standard reduction relation stuttering bisimulates the relation reified from an infinite stepping of the widened tabular semantics.
  This only holds if the state selection from the frontier is fair, since unfair choices can leave some function calls unresolved.
  Formally,
  \begin{equation*}
    \setbuild{\mtrace}{\mtrace : i \le \omega, \mstate_0 \longmapsto^i \mstate} \simeq {\mathcal F}(\mstate_0)^\omega
  \end{equation*}
  where $\simeq$ is TODO
by coinduction and appeals to the table invariants.
  If a new function result is found after a call already used the memo table, the context of the memo user is used after memoizing, so it looks like that result was part of the original memo table lookup.
  The fair selection guarantees that all function results will be eventually discovered, so the memo table appeals eventually catch up to the found results.
\end{proof}
}

@@

%%
CFA2 also includes what they called ``transitive summaries'' to deal with tail calls.
%
We create an $\rt$ continuation at every function call.
%
This view contends with tail calls, but we can identify tail calls easily --- any call with an $\rt$ as its continuation is a tail call.
%
Tail calls are important in language implementations for space complexity reasons \citep{ianjohnson:clinger:tail-calls:1998}, but in an analysis, these concerns are less important since there is no unbounded recursion.
%
The repeated popping of $\rt$ inserted by what otherwise were tail-calls is synonymous with CFA2's transitive summaries.
